---
title: "Homework 5"
author: "Benzon Carlitos Salazar (salazarbc24@uww.edu)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
options(knitr.kable.NA = "")
options(knitr.duplicate.label = "allow")
options(gt.html_tag_check = FALSE)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(magrittr)
```

# Homework for Data Visualization, Data Cleaning and Data Transformation

1. Make a histogram of the carat variable in the diamonds dataset that is available when you load the tidyverse 
package. Experiment with different binwidth values. What value reveals the most interesting pattern? Try to see if you 
separate it into different types to see a particular pattern.

```{r question_1}
diamonds_data <- ggplot2::diamonds

ggplot2::ggplot(diamonds_data, ggplot2::aes(x = carat)) +
  ggplot2::geom_histogram(binwidth = 0.1,
                          fill = "steelblue",
                          color = "black") +
  ggplot2::labs(title = "Histogram of Diamond Carat", x = "Carat", y = "Count")

ggplot2::ggplot(diamonds_data, ggplot2::aes(x = carat)) +
  ggplot2::geom_histogram(binwidth = 0.05,
                          fill = "steelblue",
                          color = "black") +
  ggplot2::labs(title = "Histogram of Diamond Carat (Binwidth = 0.05)", x = "Carat", y = "Count")

ggplot2::ggplot(diamonds_data, ggplot2::aes(x = carat)) +
  ggplot2::geom_histogram(binwidth = 0.01,
                          fill = "steelblue",
                          color = "black") +
  ggplot2::labs(title = "Histogram of Diamond Carat (Binwidth = 0.01)", x = "Carat", y = "Count")

# Facets
ggplot2::ggplot(diamonds_data, ggplot2::aes(x = carat)) +
  ggplot2::geom_histogram(binwidth = 0.05,
                          fill = "orange",
                          color = "black") +
  ggplot2::facet_wrap( ~ cut, scales = "free_y") +
  ggplot2::labs(title = "Carat Distribution by Cut", x = "Carat", y = "Count")

ggplot2::ggplot(diamonds_data, ggplot2::aes(x = carat)) +
  ggplot2::geom_histogram(binwidth = 0.05,
                          fill = "orange",
                          color = "black") +
  ggplot2::facet_wrap( ~ color, scales = "free_y") +
  ggplot2::labs(title = "Carat Distribution by Color", x = "Carat", y = "Count")
```

- A binwidth of 0.1 gives a general overview but may smooth out smaller patterns.
- A binwidth of 0.01 or 0.05 reveals discrete carat groupings (e.g., 0.30, 0.50, 1.00 carat), which reflect consumer or 
manufacturing preferences.
- When you facet by `cut`, it shows how different cuts are distributed across carat sizes. For example, Ideal cuts are 
common at lower carats, while Fair cuts may appear more in higher carat diamonds.
- When you facet by `color`, smaller weights still dominate the across all colors due to market demand. Colors I and J 
definitely shows a little bit wider carat distributions. Colors D and E shows narrower carat distributions.

2. Make a scatterplot of bill_depth_mm versus bill_length_mm and color the points by species (penguins data set). What 
does adding coloring by species reveal about the relationship between these two variables? What about faceting by species?

```{r question_2a}
ggplot2::ggplot(palmerpenguins::penguins, 
                ggplot2::aes(x = bill_length_mm, 
                             y = bill_depth_mm, 
                             color = species)) +
  ggplot2::geom_point(na.rm = TRUE) +
  ggplot2::labs(title = "Bill Depth vs. Bill Length Colored by Species", 
                x = "Bill Length (mm)", 
                y = "Bill Depth (mm)") +
  ggplot2::theme_minimal()
```

Colored by species, we get the following:

- Reveals clear groupings by species.
- Gentoo tend to have longer bills and shallower depths.
- Adelie have shorter and deeper bills.
- Chinstrap fall somewhere in between.

Coloring helps you see clusters and overlap between species in a shared plot.

```{r question_2b}
ggplot2::ggplot(palmerpenguins::penguins, 
                ggplot2::aes(x = bill_length_mm, 
                             y = bill_depth_mm)) +
  ggplot2::geom_point(ggplot2::aes(color = species), na.rm = TRUE) +
  ggplot2::facet_wrap(~species) +
  ggplot2::labs(title = "Bill Depth vs. Bill Length Faceted by Species",
                x = "Bill Length (mm)",
                y = "Bill Depth (mm)") +
  ggplot2::theme_minimal()
```

Faceting by species:

- Separates species into different panels.
- Makes within-species trends easier to analyze.
- Helps compare slope and spread of the relationship for each species without interference from others.

3. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights. Show all 
your work. (Optional Homework)

```{r question_3}
# dplyr select
flights_1 <- 
  dplyr::select(nycflights13::flights, dep_time, dep_delay, arr_time, arr_delay)

# dplyr + piping `%>%`
flights_2 <- 
  nycflights13::flights %>%
  dplyr::select(dep_time, dep_delay, arr_time, arr_delay)

# Using column indexes
flights_3 <-
  nycflights13::flights[, c("dep_time", "dep_delay", "arr_time", "arr_delay")]

# subset
flights_4 <-
  subset(nycflights13::flights, select = c(dep_time, dep_delay, arr_time, arr_delay))

# using `dplyr::select(dplyr::starts_with())`
flights_5 <-
  nycflights13::flights %>%
  dplyr::select(dplyr::starts_with("dep"), dplyr::starts_with("arr"))

# using `dplyr::select(dplyr::matches())` + regex
flights_6 <-
  nycflights13::flights %>%
  dplyr::select(dplyr::matches("^(dep|arr)_(time|delay)"))

# Base R list indexing
flights_7 <-
  data.frame(
    dep_time = nycflights13::flights[["dep_time"]],
    dep_delay = nycflights13::flights[["dep_delay"]],
    arr_time = nycflights13::flights[["arr_time"]],
    arr_delay = nycflights13::flights[["arr_delay"]]
  )

# using `with()`
flights_8 <-
  with(nycflights13::flights, data.frame(dep_time, dep_delay, arr_time, arr_delay))

# `tidyselect::one_of()`
flights_9 <-
  nycflights13::flights %>%
  dplyr::select(tidyselect::one_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
```

4. Compare dep_time, sched_dep_time and dep_delay. How would you expect those three numbers to be related?

```{r question_4a}
flights <- 
  nycflights13::flights %>%
  dplyr::select(dep_time, sched_dep_time, dep_delay)

flights$dep_time %>%
  dplyr::glimpse()
```

- `dep_time`: The actual departure time of the flight (in local time, typically in `hhmm` format).

```{r question_4b}
flights$sched_dep_time %>%
  dplyr::glimpse()
```

- `sched_dep_time`: The scheduled departure time (also in `hhmm` format).

```{r question_4c}
flights$dep_delay %>%
  dplyr::glimpse()
```

- `dep_delay`: The departure delay in minutes, calculated as: 

`dep_delay = dep_time(in minutes) - sched_dep_time(in minutes)`

You would expect that `dep_delay` is approximately 

`minutes_since_midnight(dep_time) - minutes_since_midnight(sched_dep_time)`.

- If a flight left on time, then `dep_delay=0`, and `dep_time == sched_temp_time`.
- If a flight was delayed, then `dep_time > sched_dep_time` and `dep_delay > 0`.
- If a flight left early, then `dep_time < sched_dep_time` and `dep_delay < 0`.

5. Find the flights that are most delayed upon departure from each destination. Show your R code and your solution printed.

```{r question_5}
flights <- 
  nycflights13::flights

most_delayed_by_dest <- 
  flights %>%
  dplyr::filter(!is.na(dep_delay)) %>%
  dplyr::group_by(dest) %>%
  dplyr::filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>%
  dplyr::ungroup()

most_delayed_by_dest %>%
  dplyr::select(dep_time,
                sched_dep_time,
                arr_time,
                sched_arr_time,
                carrier,
                flight,
                origin,
                dest) %>%
  dplyr::arrange(sched_dep_time) %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling(
    position = "left",
    latex_options = c("striped", "repeat_header", "scale_down"),
    stripe_color = "gray!15"
  )
```

```{r clean_up, include=FALSE}
rm(list = ls(all.names = TRUE)) # clear all objects including hidden objects
invisible(gc()) # free up memory
```
