---
title: "Homework 6"
author: "Benzon Carlitos Salazar (salazarbc24@uww.edu)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
options(knitr.kable.NA = "")
options(knitr.duplicate.label = "allow")
options(gt.html_tag_check = FALSE)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(magrittr)
```

# Hypothesis Testing

1. The mean serum – creatinine level measured in 12 patients 24 hours after they received a newly proposed antibiotic was 1.2 mg/dl.

a) If the mean and standard deviation of serum creatinine in the general population are 1.0 and 0.4 mg/dl, respectively 
then using a significance level of 0.05, test whether the mean serum creatinine level in this group is different than 
from that of the general population.

b) What is the p-value for the test?

- **Null hypothesis** ($H_0$): $\mu = 1.0$
- **Alternative hypothesis** ($H_A$): $\mu \ne 1.0$

at a significance level of $\alpha = 0.05$.

```{r problem_1}
sample_mean <- 1.2
population_mean <- 1.0
population_sd <- 0.4
n <- 12

standard_error <- population_sd / sqrt(n)
z <- (sample_mean - population_mean) / standard_error
p_value <- 2 * (1 - pnorm(abs(z)))
```

* **Z-score**: `r round(z, 3)`
* **P-value**: `r round(p_value, 4)`

Since the p-value is `r round(p_value, 4)`, which is greater than the significance level of 0.05, we **fail to reject the null hypothesis**.
There is **not sufficient evidence** to conclude that the mean serum creatinine level in the treated group is different from the general population.

2. Suppose the incidence rate of myocardial infection (MI) was 5 per 1000 among 45 to 54 year old men in 1990. To look 
at the changes in incidence over time, 5000 men in this age group were followed for 1 year starting in 2025. Fifteen new 
cases were found of MI.

a) Test the hypothesis using significance level of 0.05, that incidence rates have changed from 1990 to 2025.
b) Report a p-value corresponding to part a. 

```{r part_a_b}
observed_cases <- 15
person_years <- 5000
rate_1990 <- 5 / 1000

expected_cases <- person_years * rate_1990
poisson.test(observed_cases, T = person_years, r = rate_1990, alternative = "two.sided")
```

Suppose that 25% of patients with MI in 1990 died within 24 hours. This proportion is called the 24 – hour case fatality rate.

c) If the 15 new MI cases in the preceding study, 5 died within 24 hours, test whether the 24 – hour case fatality rate 
changed from 1990 to 2025.

```{r partc}
binom.test(x = 5, n = 15, p = 0.25, alternative = "two.sided")
```

d) Suppose we eventually plan to accumulate 50 MI cases during the period 2025-2030. Assume that the 24-hour case fatality 
rate is truly 20% during this period. How much power would such a study have in distinguishing between case-fatality rates 
in 1990 and 2025 – 2030 if a two-sided test with significance level 0.05 is planned?

```{r part_d}
# Effective Size
ES.h <- function(p1, p2) {
  2 * asin(sqrt(p1)) - 2 * asin(sqrt(p2))
}

pwr::pwr.2p.test(h = ES.h(0.25, 0.20), n = 50, sig.level = 0.05, alternative = "two.sided")
```

e) Using R software, show how large a sample is needed in part (d) to achieve 90% power?

```{r part_e}
pwr::pwr.2p.test(h = ES.h(0.25, 0.20), power = 0.90, sig.level = 0.05, alternative = "two.sided")
```

```{r clean_up, include=FALSE}
rm(list = ls(all.names = TRUE)) # clear all objects including hidden objects
invisible(gc()) # free up memory
```
